---
title: "Workshop3"
author: "Dzaki Widanto"
date: "2025-04-10"
output: html_document
---

# WORKSHOP 3
```{r setup, include=FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
library(janitor)
library(tidyverse)
library(dplyr)
library(viridis)
```
## 3.1. Tidyr

### 3.1.1. Making a tidy data
The **tidy format**:
Variables in Vertical (column).
Observations in horizontal (row).
Values in each cell!

**PRO TIP**: Always use tibble(df) to see if your data is tidy!

```{r Showing tidy data}
#library(tidyverse)
table1 # This data is tidy!
table2 # This data is untidy!
```
### 3.1.2. Pipe coding
pipe (%>% or |>) is executing multiple functions within the same variable
kinda like this:
My intended output: ABCD
A |> add (B) |> add (C) |> add (D)

```{r Using pipe feature (%>%)}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000) # Mutate adds new column

# Compute cases per year
table1 %>% 
  count(year, wt = cases) # does not add new column but instead get a new tibble

# Visualise changes over time
#library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_point(aes(colour = country)) +
  geom_line(aes(group = country), colour = "grey50") 
  
```
## 3.2. Pivoting dataset

Most of the data that we would process will be untidy.
Especially, when:
a. one variable spread across multiple columns
b. One observation spread across multiple rows

Solution: **pivot** to lengthen or widen the data!!!

### 3.2.1. Pivot - Lengthening
```{r}
# billboard # The data is still wide, contains wk1 - wk76
```
Here, each observation (row) is dedicated for song (track)
The columns (wk1-wk76) describe variables of the song!
However, these columns contains two meaning!
a. variable (week)
b. values (rank, number)

Therefore, we should separate the variable week with the value rank into individual columns!
### 3.2.2. Pivoting process
```{r data lengthening - pivot}
billboard |>
  pivot_longer(
    cols = starts_with("wk"), # cols: columns to pivot into longer format, becomes ROWS and variable
    names_to = "week", # Character vector, change the name; naming the header of new column in cols
    values_to = "rank", # specifying the name of column to create from the data stored in cell values
    values_drop_na = TRUE # this line is to delete rows with NA!
  )
# keeping the values (rank number) identical to each cols (week) of observation (track)

# These code lines are also act the same.
  
#billboard |>
#  pivot_longer(
#    cols = !c(artist, track, date.entered), # same as starts_with
    # this is more useful if the column names are very different
#    names_to = "week", 
#    values_to = "rank" 
#  )
```
When we pivot the column names, the column that does not have values **automatically turns into NA**.
(example: 2 Pac Baby dont cry wk8 - wk10)

### 3.2.3. Pivoting longer

Examine how pivot does to our data
```{r Pivot lengthening examination}

# create a tiblbe
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
# pivoting
# to make bp1 and bp2 as new column, whilst creating one observation = one row!
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```

### 3.2.5. Widening datasets
```{r}
# Example data
cms_patient_experience # Already tidy!

# Examine unique values in measure_cd and measure_title
cms_patient_experience |> 
  distinct(measure_cd, measure_title)

# Data pivot - Widening
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"), # The columns that are staying
    names_from = measure_cd, # widen the measure_cd 
    values_from = prf_rate # make separate columns for different values (prf_rate) of observations (measure_cd)
  )
# Now each org_.. has individual values,
# and the variable measure_cd is widened into CAHPS_GRP_x variables
# with the values from prf_rate
```
### 3.2.6 Widening Longer
Examine how pivot_wider does to our data
```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )

# What pivot_wider() does in a nutshell
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
```
To start pivoting process, pivot_wider() needs to figure out:
a. Column names --> Unique values in measurement (example)
b. Rows --> from column that will not be changed, aka id (example)

### 3.2.7. Pivot - Exercise 1
1.	Why are pivot_longer() and pivot_wider() not perfectly symmetrical? 
Carefully consider the following example
```{r}
# Initial data
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(1,2,1,2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
# Step 1: pivot wider
stocks %>% 
  pivot_wider(
    names_from = year,
    values_from = return) %>% # This one works, it widens the tibble
  # Step 2: pivot longer
  pivot_longer(cols = `2015`:`2016`, names_to = "year", values_to = "return")
```
Here, we see that in the first stocks tibble,
year (dbl), half(dbl), return(dbl)

Then, when we pivot_wider and pivot_longer, the final result is half(dbl), year(chr), return(dbl).

See that the year becomes character. This is because when we pivot_wider, 2015 and 2016
in the column name becomes character!

Then when we pivot_longer to return it back as a variable (column), it stores it as character.

Now, let's try to fix this!
```{r}
# Then, we fix the data type with:
stocks %>% 
  pivot_wider(
    names_from = year,
    values_from = return) %>% 
  pivot_longer(`2015`:`2016`,
    names_to = "year", 
    names_transform = list(year = as.numeric), # transform the year column from character into numeric
    values_to = "return")
```

### 3.2.8. Exercise 2
Why this code fails?
```{r}
table4a %>% 
  pivot_longer(
    c(1999, 2000),
    names_to = "year", 
    values_to = "cases")
```
Here, the declaration is not correct.
remember, as a column name, 1999 and 2000 are characters!
therefore, we need to declare them as characters not numbers.

```{r Exercise 2 solution}
# Change into
table4a %>% 
  pivot_longer(
    c("1999", "2000"), # Declare the column names as characters
    names_to = "year",
     names_transform = list(year = as.numeric),
    values_to = "cases")
# Now, the data is tidy!
```

### 3.2.9. Exercise 3
3.	Consider the sample tibble below. Do you need to make it wider or longer? 
What are the variables?
```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
```

Solution: Pivot lengthening the data!
```{r Exercise 3 solution}

# Need to lengthen pivot!
preg %>%
  pivot_longer(
    cols = contains("male"),
    names_to = "gender",
    values_to = "count"
  )
```
## 3.3. Separating and uniting data tables in tibble

### 3.3.1. Separate()

separate() in default, separates a value when it detects non-number or letter character!
```{r}
# Separate can also detect numbers. We just need to declare if the separated values
# needs to be the natural type (for example, number will be converted into double!)
table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE) 
# convert = T is to make cases and population values double (their natural type)

# Other type of separation, based on number of characters!
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```
### Unite()
```{r}
table5 %>% 
  unite(new, century, year, sep = "")
# Make a new column "new"
# Then, unite century and year
# with separators "" (none), so it concacenate together!
```

## 3.4. Handling Missing Values

### 3.4.1. Explicit missing value - NA
NA is stated!
```{r}
treatment <- tribble(
  ~person,           ~treatment, ~response,
  "Derrick Whitmore", 1,         7,
  NA,                 2,         10,
  NA,                 3,         NA,
  "Katherine Burke",  1,         4
) 
treatment |>
  fill(everything())
# last observation carried forward, aka last available values is being copied to the next NA rows
```
### 3.4.2. Explicit missing value - Fixed Value
if its form of fixed value (e.g.: -99 or -9999)

```{r}
x <- c(1, 4, 5, 7, NA)
coalesce(x, 0)  # Change NA to 0
#> [1] 1 4 5 7 0

x <- c(1, 4, 5, 7, -99)
na_if(x, -99)  # Conditional, change the stated fixed value into NA
#> [1]  1  4  5  7 NA
```
### 3.4.3. Explicit missing value - NaN (Not a Number)
typically behaves **same as NaN**

```{r}
x <- c(NA, NaN)
x * 10
#> [1]  NA NaN
x == 1
#> [1] NA NA
is.na(x)
#> [1] TRUE TRUE
```
### 3.4.4. Implicit missing data
```{r}
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
# missing data:
# a. price in year 2020 qtr 4
# b. completely unwritten, year 2021 qtr 1

stocks |>
  pivot_wider(
    names_from = qtr, 
    values_from = price
  ) # The missing year 2021 qtr 1 is shown


```
## 3.5. Relational Data
Simply put, joining dataframes together!

Main package: dplyr.

3 families of relational data:
a. Mutating joins = add new variables to one dataframe from matching observations
b. filtering joins = filter observations from one data frame based on whether
or not they match an observation in the other table
c. Set operation = treat observations as if they are set elements

### 3.5.1. Joining Datasets
Join datasets by identifying the **keys**.

**Keys**: A variable/set of variables that uniquely identifies an observation.
in SC5502, keys are the data ID (unique per observation, only 1!)

if the keys are the unique, main identity of a dataframe --> **primary**
if the keys are unique, but NOT main identity of dataframe --> **foreign**
```{r}
library(nycflights13)

airlines # First data
airports # second data
planes # third data. Primary key is tailnum
flights # Fourth data, THE MAIN DATA (COMPLETE).
# in Flights, tailnum is unique identity but only Foreign key
```
### 3.5.2. Identifying primary keys in datasets
filter n>1 result is not visible (because the variables only have 1 observation each!)

if data does not have primary key --> create one by mutate!!!

Primary key and corresponding foreign key forms relation!
```{r}
planes %>% 
  count(tailnum) %>% 
  filter(n > 1)  #tailnum is the primary key

weather %>% 
  count(year, month, day, hour, origin) %>% 
  filter(n > 1) 
weather |>
  count(origin) |>
  filter(n>1)

# Data that does not have primary key
flights %>% 
  count(year, month, day, flight) %>% 
  filter(n > 1) # Still has results, mainly n =2 or n = 3. NO clear unique identity

flights %>% 
  count(year, month, day, tailnum) %>% 
  filter(n > 1)
```
### 3.5.3. Mutating Joins
Combine variables from two tables.
adds to the **right side** of the data!!! Matches by row (observation) and corresponding keys!

```{r}
flights2 <- flights %>% 
  select(year:day, hour, origin, dest, tailnum, carrier)

airlines # has 2 variables: carrier, name

# flights2 >< airlines via key CARRIER

flights2 %>%
  select(-origin, -dest) %>%  # drop down origin and dest
  left_join(airlines, by = "carrier") # the flights2 data is joined to the left side of airlines

# A way more complicated version
flights2 %>%
  select(-origin, -dest) %>% 
  mutate(name = airlines$name[match(carrier, airlines$carrier)])
```
### 3.5.4. Inner join
```{r Visual of Inner join}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)
x %>% 
  inner_join(y, by = "key") # deleted 3 and 4 in key
```

### 3.5.5. Outer join
left_join() --> keep observation in x
right_join() --> keep all observations in y
full_join() --> keep all observations in x and y

All inexistent data will be considered as NA

### 3.5.6. Join with duplicate keys (non-unique)
```{r duplicate keys}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     1, "x4"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2"
)
left_join(x, y, by = "key") # create duplicates of 2 and 1, fits all
```

```{r}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     3, "x4"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     2, "y3",
     3, "y4"
)
left_join(x, y, by = "key") # make duplicates in x (x2 and x3)

      # Default, by = NULL - Joins all!
flights2 %>%  # year:hour, origin, dest, tailnum, carrier
  left_join(weather) # weather: origin, year:hour, temp, ...
# output: year:hour, origin, dest, tailnum, carrier, temp, ...
# here the keys are year:hour, origin.

      # Join by character vector
flights2 %>% # year:hour, origin, dest, tailnum, carrier
  left_join(planes, by = "tailnum") # tailnum, year, type, ...
# results: year.x, month:hour, origin, dest, TAILNUM, carrier, year.y

# Here, year in flights2 and tailnum means different!
# It separate the variables (columns)

      # join by NAMED character vector by = c("a" = "b")
# Match variable a in table x to variable b in table y

flights2 %>% # year:hour, origin, dest, tailnum, carrier
  left_join(airports, c("dest" = "faa")) # faa, name, lat, lon, ....
# output: year:hour, origin, dest, tailnum, carrier, name, lat, lon
# here, the faa in airports is considered the merging key of dest!

flights2 %>% 
  left_join(airports, c("origin" = "faa")) # joining flights2 with airports by 
# matching the origin with faa. 

```