---
title: "Workshop4"
author: "Dzaki Widanto"
date: "2025-04-10"
output: html_document
---
# 4.1. Packages and Data
```{r setup, include=FALSE, message = FALSE, warning = F}
#install.packages("sf")
#install.packages("tmap")
#install.packages("mgcv")
library(ggplot2)
library(janitor)
library(tidyverse)
library(dplyr)
library(terra)
library(sf)
library(tmap)
library(mgcv)
library(readr)
```

## 4.1.1. Data
```{r}
#load the copepod data into R studio
dat <- read_csv("../data/data-for-course/copepods_raw.csv")
dat

# Check primary keys
dat %>%
  count(silk_id, segment_no) %>% 
  filter(n > 1)  # silk_id + segment_no are the primary key
```
# 4.2. Data Exploration
Entering the longitude and latitude in ggplot.

however, this **IS NOT MAP**, despite looking like it.
```{r load coordinates}
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()

# Plot species richness against latitutde
ggplot(dat, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()

```
# 4.3. Exploring the Map tools

in **sf** package:
st_as_sf = converts different data types to simple features
crs = coordinate reference system

Turns out, crs4326 is WGS 84 System!
```{r get going with maps}
sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), # this function makes new column, geometry
                 crs = 4326) #st_as_sf converts into simple feature
tibble(sdat)

# Checkout the crs meaning
crs4326 <- st_crs(4326)
crs4326 # look at the whole CRS
crs4326$Name # pull out just the name of the crs
# crs4326$wkt 

```
# 4.4. Feature Collection
simple feature ~ shapefile! 
so in one column it can hold multiple data, such as latitute and longitude, as a single POINT
so simple feature can also store things such as polygon!!!
```{r}
sdat
plot(sdat["richness_raw"]) # eventhough its only richness, it already has POINT data
# Therefore it pulls up like in the map

plot(sdat) # This function plots all variables (column) based on geometry (point)
```
# 4.5. Plotting maps
Here, we gonna use **tmap** package!
```{r}
#using tmap

tm1 <- tm_shape(sdat) + # if only run this, shows nothing
  tm_dots(col = "richness_raw", palette = "viridis")    # plot column richness_raw

tmap_save(tm1, filename = "../output/Richness-map.png", 
          width = 7, height = 5, units = "in")

# saving the plot
```

# 4.6. Mapping spatial polygons as layers
## 4.6.1. Loading data (shapefile)
Note: Shapefile is inefficient bcs (if you remember) it has multiple files! easily lost!

Alternative is **geopackage** (.gpkg)
```{r}
# Example using shapefile
aus <- st_read("../data/data-for-course/spatial-data/Aussie/Aussie.shp")

# Data contains polygon of Aussie states

shelf <- st_read("../data/data-for-course/spatial-data/aus_shelf/aus_shelf.shp")
aus
```
## 4.6.2. Mapping my polygons
tm_shape(shp, bbox)
bbox: Bounding box of the map

1. tm_shape = plot multipoints (MULTIPOLYGON)
2. tm_dots = plot singular point data (POINTS)
3. tmap_style = similar to ggplot themes()
```{r ggplot but tmap!}
# Plotting the polygons
tm2 <- tm_shape(shelf) + 
  tm_polygons()
tm2

# we can modify this further as in ggplot!!!
# in mapping, we can make layers of polygons and data!

tm_shape(shelf, bbox = sdat) +  # shelf is base polygons, bbox is the boundary of the map
  tm_polygons() + # Plot the polygons
  tm_shape(aus) + # We layer the aus shp (borders)
  tm_polygons() + # and we plot the aus shp borders
  tm_shape(sdat) + # we plot our data!
  tm_dots()    +   # to plot the dots!
  tmap_style("cobalt") # To style your map, kinda like color scale
```
# 4.7. Further exploration of t_map
```{r}
library(tmap)

# load data
data("World")

tm_shape(World) +
  tm_polygons() +
  tm_crs(crs = 4326) +
  tm_polygons("pop_est", palette = "viridis", title = "Population") +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(legend.outside = TRUE) +
  tm_text("name", size = 0.4) +
  tmap_style("classic")
```
## 4.7.1.  Make Interactive map
```{r}
tmap_mode("view")  # Make into viewer mode!

my_tmap <- tm_shape(World) +
  tm_polygons("pop_est", palette = "viridis", title = "Population")
my_tmap
tmap_save(my_tmap, filename = "HtmlWidget/416-introduction-to-tmap-interactive.html")
# Error mungkin ga bisa

tmap_mode("plot")  # to return into plot map
```


